\pagestyle{fancyplain}
\fancyhf{}
\fancyhead[R]{\nouppercase\chaptername \space \thechapter}
\fancyfoot[R]{\thepage}
\pagenumbering{arabic}

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = FALSE,message = FALSE,error = FALSE, warning = FALSE, 
                      fig.width = 16.6/2.54, fig.height = 10/2.54, dpi = 800, 
                      fig.path = "./output/")
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", 
               "#0072B2", "#D55E00", "#CC79A7")
```

# General introduction

\begin{figure}[!htb]
\centering
\includegraphics[width=15cm]{pics/Rainforest1.jpg}
\caption*{Sunshine through rainforest canopy in Danum Valley.}
\end{figure}

\pagebreak

## Threats to biodiversity

Throughout the Anthropocene, humans have faced crises. In 2000 the United Nations developed eight goals for 2015, known as the Millennium Development Goals, one of which was to 'ensure environmental sustainability' [@united_nations_united2014]. Amongst other things, this goal is in recognition of the current extinction crisis. Recent extinction rates far exceed their pre-human levels [@pimm_future1995], and are close to constituting the 6th mass extinction event [@barnosky_has2011].

Humans are at heart of the extinction crisis, but which of our environmental impacts is principally to blame? Five key threats are: land-use change, climate change, pollution, over-exploitation and invasive species [@hirsch_global2010]. Whilst the greatest overall threat to terrestrial systems is currently land-use change, climate change is forecast to become increasingly important [@sala_global2000]. 

Having diagnosed the threats for biodiversity, we cannot assuage them until we identify underlying drivers. Climate change is driven by changes in: (1) atmospheric concentrations of greenhouse gases (GHGs) and aerosols, (2) land cover and (3) solar radiation [@ipcc2013]. All of these changes occur naturally, but climate change since pre-industrial times is primarily caused by anthropogenic emissions of GHGs from the burning of fossil fuels and through land-use change [@ipcc2013]. Land-use change includes both wholesale conversion and degradation. Generally habitat is converted to create agricultural land to feed the growing human population [@foley_solutions2011; @godfray_food2010]. Degradation of remnant, unconverted habitat may result through incipient fragmentation. Additionally, habitat degradation is caused by selective logging, hunting and fire – the key is that the overall habitat type remains the same but the quality declines. 

Given the importance of the underlying drivers of climate change and land-use change for the persistence of the human population, it is unrealistic to expect these pressures to cease. One option is to mitigate change by stemming human population growth and increasing the efficiency of resource acquisition [@godfray_food2010]. 

Alternatively, the biodiversity crisis could be alleviated through a better understanding of how and why organisms respond to human impacts. In this way, we could modify our actions to minimise impact, and also facilitate organism responses that permit persistence through change. This is the step that I will address in the following review. Initially I will focus on organism responses to climate change, given the increasing importance of this pressure in the future [@sala_global2000]. However, neither the impacts of climate change nor land-use change can be fully understood in isolation; the synergies between the two pressures are thought to be extensive, but generally poorly understood [@brodie_climate2012; @mantyka-pringle_interactions2012]. In the tropics forest degradation is some 20 times more pervasive than deforestation [@asner_contemporary2009], yet there is particularly little discussion of how habitat degradation might interact with climate change. This is the key unexplored area that I will move on to discuss, before finally outlining my PhD framework. 

## Responses to climate change

There are three possible outcomes for organisms experiencing environmental change: (1) they die, (2) they move to more optimal environmental conditions, or (3) they adapt in situ to the new environmental conditions. The first case results where organisms fail to adequately implement either of the latter two adaptive responses to change. 

### Extinctions due to climate change

A species is classed as extinct on the IUCN Red List if "there is no reasonable doubt that the last individual has died" [@baillie2004]. Twenty-five species are classified as extinct or extinct in the wild owing partially or wholly to "climate change and severe weather" [@iucn_iucn2014]. Between 1880 and 2012, global average temperature increased by 0.85&deg;. This trend will continue into the future, with predictions of global average temperature for the period 2081-2100, relative to 1986-2005, ranging from an increase of 1-3.7&deg;C, depending on the scenario used [@ipcc2013]. 

Evidently the increase in global average temperature occurs on a long timescale and in concert with many other human impacts, so it can be difficult to directly attribute biodiversity loss to this change per se. The most obvious proximate cause of extinction directly due to increasing average temperature is loss of climatically-suitable habitat [@thomas_extinction2004], but examples under current climate change have yet to manifest.

Where extinctions have been attributed to climate change, this is through changes in local weather patterns. Weather is distinguished from climate as being "the state of the atmosphere at a given time and place", whereas climate comprises "the statistics of weather conditions over a decade or more" [@ipcc2013]. Concomitant with increasing global average temperature is the increase in the frequency and intensity of extreme weather events [@ipcc2013]. This can be explained statistically, because an 'extreme weather event' is an event in which the climatic conditions fall towards either extreme end of the probability distribution (the 10th or 90th percentile; @ipcc2013]. Provided the probability distribution of temperatures remains the same (or similar), an increase in average temperature corresponds to an upwards shift in the overall temperature distribution, and therefore we more commonly see temperatures that were originally very rare, and begin to see temperatures never before recorded (\autoref{fig:fig-1-1}). It is almost certain that there will be more extremes of heat (and fewer extremes of cold) towards the late 21st Century [@ipcc2013]. 

Future changes in precipitation are more difficult to predict than changes in local temperature, but precipitation events also play a significant role in species' extinctions due to climate change. For example, extremely hot and dry years significantly contributed to the extinction of the golden toad [@pounds_biological1999]. It is likely that that heavy precipitation events will increase in frequency and/or intensity over many land areas, whilst the intensity and/or duration of droughts may also increase towards the late 21st Century [@ipcc2013].

```{r fig-1-1}
```
\begin{figure}
\centering
\includegraphics{figs/Fig1.1.jpg}
\caption{(\#fig:fig-1-1)Schematic showing the increase in frequency of extreme
temperatures (shaded light pink) and the magnitude of extreme
temperatures (shaded dark pink), in response to increasing mean
temperature for a normal distribution of temperatures. `extreme' refers
to events that would have been anomalous under the previous probability
distribution. Figure taken directly from \citet{ipcc_climate2007}.}
\end{figure}

### Range shifts due to climate change

Species may track optimal climatic conditions by shifting their range. This commonly occurs through net population extinctions at the trailing edge, or net population colonisations at the leading edge [@parmesan_poleward1999]. Dispersal by individuals may also occur in highly mobile species. Since the predominant effect of climate change is increasing temperature, many species track temperature by moving to higher latitudes -– as exemplified in the Arctic, where organisms such as shrubs and red foxes have expanded polewards [@hersteinsson_interspecific1992; @sturm_climate2001]. Others move to higher altitudes; both latitudinal and altitudinal shifts have been seen in birds and butterflies of temperate regions [@hill_responses2002; @parmesan_poleward1999; @thomas_birds1999]. 

Until recently, there were very few studies of range shifts due to climate change in tropical species, with some suggesting that the response should be less extreme given the slower rates of warming in the tropics [@freeman_rapid2014; @ipcc2013]. It is now apparent that tropical species do shift their ranges to track climate, particularly to higher elevations [@chen_elevation2009; @pounds_biological1999] owing to shallow temperature gradients across latitudes [@colwell_global2008]. In fact, tropical species track climate more closely than temperate species [@freeman_rapid2014]. This effect could be due to: (1) greater thermal specialisation as a result of long-term thermal stability in the tropics [@freeman_rapid2014]; (2) slower velocity of climate change up mountains [@loarie_velocity2009] meaning it is easier for species to keep pace; or (3) fewer barriers to dispersal in the tropics, since tropical biomes have thus far retained a greater proportion of natural habitat than temperate regions. In any case, even tropical species do not track climate precisely [@chen_elevation2009].

### In situ adaptation to climate change

In situ adaptation encompasses biochemical buffering, gene expression, phenotypic plasticity, behaviour and genetic adaptation [@peck_organisms2011]. Adaptation is complex and largely unpredictable; hence it is rarely accounted for in models used to predict range shifts [@peck_organisms2011]. This may be one of the reasons that species do not move as quickly as predicted.

Modifications in species' phenology represent the vast majority of documented adaptations to climate change in situ. Many of these examples come from temperate regions of the Northern hemisphere, where seasonality is the overarching determinant of species' phenology, and is itself dramatically altered by climate change [@bradshaw_evolutionary2006]. Specifically, spring has advanced and the growing season has lengthened. Organism responses include earlier breeding in animals such as birds and butterflies, earlier arrival of migratory birds, and earlier flowering in plants [@walther_ecological2002]. 

Responses such as physiological plasticity or genetic adaptation feature much less in the literature on in situ adaptation to climate change. This may signify a real scarcity of such changes in nature. The evolution of new forms that enable persistence in the same geographic range under a changing climate requires a species to become tolerant of a climatic regime to which it was previously intolerant, which seems unlikely [@parmesan_ecological2006]. One could argue that selective pressures to evolve increased thermal tolerance would have been insufficient prior to present-day climate change. However, major evolution at the species level is not evident in the fossil record during the Pleistocene glaciation event, even though this comprised climate change of 5-10 times the magnitude of 20th Century warming [@parmesan_ecological2006].

While evidence for evolutionary responses to climate change is limited, this is not to say that evolution has no role to play. Where examples of such responses do exist, they underlie aforementioned ecological changes in phenology or dispersal [@parmesan_ecological2006]. For example, Dutch great tits that display greater plasticity in their timing of reproduction are better able to match egg-laying to food availability – the peak of which has advanced as a result of climate change – and thus achieve greater fitness [@nussey_selection2005]. There are also practical explanations for the lack of documented evolutionary responses, since these adaptations are less intuitive and harder to document than ecological responses [@oconnor_toward2012]. 

The literature discussed above fails to mention one additional and very significant tool that animals can employ to adapt to climate change – behaviour. All organisms ordinarily experience a range of temperatures, and so possess thermoregulatory behaviours that can also be deployed to mitigate the impacts of climate change. Chamois, for example, move to higher altitudes and reduce activity when temperature increases [@mason_predicting2014]. 

Habitats present a considerable degree of variation in microclimates, because of variation in microhabitat features [@scheffers_microhabitats2014], slope and aspect [@suggitt_habitat2011]. Behavioural plasticity allows animals to move into these microclimates [@scheffers_microhabitats2014] and so track their optimal climate on a local scale. These so-called "microrefugia" are utilised by a variety of taxa around the world. In boreal forests of Finland, moose seek out the cooler microclimates of forests with higher and denser canopies, in response to high daytime temperatures [@melin_moose2014]. Similarly, in the tropics, possums choose the coolest tree hollows in which to den [@isaac_microclimate2008], and herpetofauna of Singapore occupy microrefugia that not only largely avoid their critical thermal maxima (CT~max~) – which is often exceeded in the wider macroclimate – but their microrefugia also heat less quickly than the macroclimate [@scheffers_microhabitats2014].

## Influence of land-use change

The most well-known interaction between climate change and land-use change is probably that the latter can cause the former, on a global scale, through the release of GHGs [@ipcc2013]. Deforestation marginally reduces the net radiative forcing that leads to global climate warming through decreases in surface albedo [@ipcc2013]. Climate change could also cause land-use change, such as through shifting the areas which are most climatically suited for agriculture [@opdam_climate2004].

In this review, however, I have focused on how organisms can adaptively respond to change. The question then becomes: how does land-use change influence an organism's capacity to respond to climate change? On a regional scale under wholesale conversion, the answer is relatively well-discussed. Namely, regional habitat loss creates barriers to climate-driven range shifts [@thomas_protected2012]. Recall that in response to climate change that has already happened, organisms have not moved as quickly as expected [@chen_elevation2009], and barriers to dispersal may contribute to this. 

In situ adaptation may allow organisms to persist in habitats from which they are unable to move, or it may remove the need to move altogether; in either case, the influence of land-use change on in situ adaptation to climate change has not been elucidated. Given that many barriers to dispersal have already been introduced, and this will likely continue into the future, it is vital to facilitate adaptation to future climate change within the areas that species already occupy.

Most obviously, wholesale conversion of natural or semi-natural habitat appears to increase local daytime mean temperature [e.g. @wickham_comparison2012]. Largely, this is caused by an increase in daily maximum temperature as a result of decreased interception by overhead vegetation of direct solar radiation [@xu_scale-dependent2004]. The effect is reversed at night when outgoing long-wave radiation is lost because of reduced interception by vegetation [@xu_scale-dependent2004]. An increase in mean temperature may exceed an organism's preferred body temperature, and so potentially lead to sublethal effects [@du_plessis_costs2012], but increasing maximum temperature generally poses the greater threat to organisms. Organisms can often acclimate to moderate increases in average temperatures [@peck_animal2009], but if their critical thermal maximum is exceeded – even for a very short amount of time – this will cause death. Thus, species remaining in habitat after it has been converted are already likely to be under some amount of thermal stress, and future climate change may push temperatures beyond the range that they can tolerate through physiological plasticity. 

Any increase in ambient temperature will ultimately increase the temperature of microclimates, and so potentially decrease their efficacy as thermal microrefugia for thermally stressed individuals. The extent to which microclimate utility is compromised depends upon the rate at which they warm alongside macroclimate warming. There is evidence from the tropics that this relationship is non-uniform, with microhabitat temperatures increasing only 0.11–0.66&deg;C for every 1&deg;C in the macroenvironment [@scheffers_microhabitats2014]. Asymmetry in warming rates will be influenced by factors that act to create the microclimate, such as the microhabitat [@scheffers_microhabitats2014], slope, aspect or elevation [@suggitt_habitat2011]. It is possible that microclimates could be entirely removed as a consequence of microhabitat removal [e.g. loss of some bird's nest fern species upon conversion of forest to oil palm plantation; @fayle_effect2009] or extreme macroclimate warming [@caillon_warming2014].

Wholesale conversion is very likely to impede the ability of persisting organisms to adapt to future climate change, but then few of the original species do persist through conversion [e.g. @gibson_primary2011; @katovai_understory2012; @murphy_meta-analysis2014] and –- at least in the tropics -– habitat degradation is far more pervasive. In particular, some 20% of the humid tropical biome experienced selective logging from 2000-2005 [@asner_contemporary2009], whilst deforestation affected only 1.4% in the same period [@hansen_humid2008]. Although the habitat type broadly remains as 'forest', selective logging can be extremely disruptive. Indeed the term 'selective' is somewhat of a misnomer, meaning that particular species and stems (usually above a minimum trunk diameter) are targeted [@edwards_maintaining2014]. These targets are typically the largest, oldest trees, the removal of which reduces canopy height and canopy density [@kumar_effects2005; @okuda_effect2003], and also fragments the forest canopy and opens up large gaps [@edwards_maintaining2014] that are often invaded by non-tree species, such as climbers and bamboo. Commercial selective logging also causes collateral damage, particularly where trees are connected by climbers [@schnitzer_recruitment2004], as well as requiring roads and skid trails that bring further challenges for wildlife [@brodie_correlation2014; @laurance_global2014], and heavy machinery that result in soil compaction [@putz_reduced-impact2008].

Since selective logging reduces canopy cover, just as deforestation does, so it is likely that the thermal regimes of degraded forest will be similarly altered. Moreover, there is already some indication that previously identified tropical microrefugia [in this case, leaf litter and soil; @scheffers_microhabitats2014], are reduced by logging [@saner_reduced2009]. Conversely, ground vegetation -– another microrefugium [@scheffers_microhabitats2014] -– may be favoured by the release of pioneer species upon the creation of treefall gaps. 

The impact of habitat degradation on species' ability to persist under climate change is likely to be less profound than under wholesale conversion, simply because the amount of habitat change is less. However, a greater proportion of species found in undisturbed habitat remain in degraded habitat than in converted habitat [@edwards_degraded2011], and it is these species that are of primary conservation concern. Furthermore, degraded forests now represent a significant proportion of the humid tropical biome, and are therefore home to a significant proportion of all tropical forest species on Earth. The potential for these species to track climate change through dispersal is limited – there are barriers, such as hostile land-use types, as well as a shallow latitudinal temperature gradient [@colwell_global2008] and a potential lack of connected, higher elevation habitat [@scriven_protected2015]. Many tropical species will need to adapt to climate change within degraded forest, if they are to persist into the future. Therefore, although we do not yet fully understand the impact of any land-use change on the ability of tropical species to adapt in situ to climate change, I argue that we should first explore the impacts of habitat degradation, as a priority. 

## Thesis aims and rationale

### Definitions

'Microhabitats' are fine-scale (mm to cm) features within a habitat, including leaf litter, deadwood, tree holes and epiphytes within rainforest habitats. Each of these features will have its own 'microclimate' which may be different from the macroclimate that acts at the level of the whole habitat (m to ha). When microhabitat features offer a more desirable microclimate than the macroclimate, the features can be referred to as 'thermal microrefugia' ('microrefugia' henceforth).

### Chapter 2 -- A pantropical analysis of the impacts of forest degradation and conversion on local temperature

SUMMARY

### Chapter 3 -- A framework for quantifying fine-scale thermal heterogeneity in the field

SUMMARY

### Chapter 4 -- Tropical forests are thermally buffered despite intensive selective logging

SUMMARY

### Chapter 5 -- The impact of recent forest cover change on climate connectivity in the tropics

SUMMARY


